---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
    
---

## Loading and preprocessing the data
In order to analyse the activity monitoring data we must first load the dataset.
In order to do this we must create a connection to the zip file containing
the data:
```{r}
connection<- unzip("activity.zip")
```

The data can be read in using the data.table package.
```{r}
library(data.table)
amd<- fread(connection)
```

We shall convert the date into the POSIXct format for convenience:
```{r, results='hide'}
# Convert date values 
amd[, date := as.POSIXct(date,format = "%Y-%m-%d")]
```

## What is mean total number of steps taken per day?
The total number of steps (excluding missing values) taken each day during the 
two months can be described with a histogram:
```{r,  results='hide'}
total_steps_per_day<- amd[!is.na(steps),sum(steps) ,by=date]
total_steps_per_day[,hist(V1,
                          breaks = 30,
                          col="red",
                          main = "",
                          xlab = "Total steps taken in a day")]
```

A summary of the the total number of steps (excluding missing values) over all
the days shows the mean number of steps in a day to be
10770 and the median to be 10760:
```{r}
summary(total_steps_per_day[,V1])
```
## What is the average daily activity pattern?
The average daily activity pattern (excluding missing values) can be seen in the
graph below where interval values range from 0-2400:

```{r, results= 'hide'}
steps_by_interval<- amd[!is.na(steps),mean(steps),by=interval]
setnames(steps_by_interval,"V1","mean_steps_taken")
steps_by_interval[,plot(interval,mean_steps_taken, type = "l")]
```

The interval during which the maximum average steps are taken starts at 8:35 a.m. 

```{r}
steps_by_interval[which.max(mean_steps_taken)]
```
## Imputing missing values
No missing values occur for either the date or interval labels:
```{r}
amd[is.na(date),sum(.N)]
amd[is.na(interval),sum(.N)]
```
The total number of missing values for the recorded steps is:
```{r}
amd[is.na(steps),sum(.N)]
```
The values will be imputed to reduce the bias introduced by omitting these
values. For each interval that was not reported, the average activity over that
interval is instead used:
```{r}
#Generate the mean for each interval
amd[,mean_steps := as.integer(mean(na.omit(steps))),by= interval]
# Created a new column for the new data with imputed values
amd[,imputed_steps:=steps]
# Assign an intervals mean value in all cases of missing data
amd[is.na(imputed_steps),imputed_steps:= as.integer(round(mean_steps))]
```

In order to assess the effect of imputing the previous analysis is carried out
once more:
The total number of steps taken each day during the 
two months can be described with a histogram:
```{r,  results='hide'}
total_steps_per_day<- amd[!is.na(imputed_steps),sum(imputed_steps) ,by=date]
total_steps_per_day[,hist(V1,
                          breaks = 30,
                          col="red",
                          main = "",
                          xlab = "Total steps taken in a day")]
```

A summary of the the total number of steps over all
the days shows the mean number of steps in a day to be
10750 and the median to be 10640:
```{r}
summary(total_steps_per_day[,V1])
```
These results are similar to before with the exception of the substantial and 
artificial increase in one bin of the histogram. Imputing missing values using
this method is not ideal.


## Are there differences in activity patterns between weekdays and weekends?
During the weekdays activity in the morning is starts earlier, persists for a
longer duration and is more intense than on the weekends. At weekends, however, activity
levels of activity are maintained more consistently over the full course of the day:
```{r, results = "hide"}
amd[,weekend_status := "weekday"]
amd[weekdays(date) %in% c("Saturday","Sunday"), weekend_status := "weekend"]
comparison_of_days<- amd[,mean(imputed_steps) ,by = .(interval, weekend_status)]
library(lattice)
comparison_of_days[,xyplot(V1 ~ interval | weekend_status,
                           type = "l" , 
                           ylab = "Mean activity (step count)",
                           layout = c(1,2))]

```
